Compositional Quantitative Resource Analysis of C Programs

### 2014-06-16 paper/main.tex
work on syntax and semantics

Let us now start the formal part of this work.  We begin by defining
a proper syntax for the language we will be using later on.  This is
a more algebraic version of the surface syntax used in the previous
examples.
%
\begin{mathpar}

S := \code{skip}
\mid \code{assert}(C)
\mid x \gets E
\mid \code{while}(C)~S
\mid \code{if}(C)~S~\code{else}~S
\mid S;S

\and

E := x \mid x \pm y
\end{mathpar}

Note that expressions are kept extremely simple, this will help
greatly when designing the analysis.  Moreover, programs
containing fancier expressions can easily be compiled to this
restricted form.  For example $a \gets b + c - d$ is semantically
equivalent to $a \gets b; a \gets a + c; a \gets a - d$ which
fits in the above syntax.

The syntax of conditions, noted $C$, is left unspecified
for the moment.  We will describe it later when the practical
implementation details are provided.  The only semantic tool
we need to deal with this unknown is an evaluation
function $\llbracket \_ \rrbracket_{\_}$ such that, for any
condition $C$ and any heap $H$ we have $\llbracket
C \rrbracket_H \in \{ \top, \bot \}$.

Another peculiarity is that there are no integer constants in the
syntax.  Once again this is for the sake of simplicity.  In practice
the concrete syntax recognizes ``variables'' like {\tt -4} or
{\tt 60} as read only variables.  Any heap used later in the
operational semantics will map these variables to their natural
value.

\section{Cost Semantics}

The semantics for the syntax defined above needs to account
for resource consumption.  To account for all kinds of resources
we leave it as a parameter of the semantics to be specified
by the user.  This parameter, called a metric, is a tuple of rational
numbers with the following form
$$
(
M_k, M_s, M_a,
M_w^1, M_w^2, M_w^3,
M_c^1, M_c^2,
M_s^1, M_s^2
).
$$
Each of these rational numbers indicates the amount of resource
consumed by an operation.  If one of these numbers is negative,
it means that some resource is released by the program.

The general form of an evaluation judgement is
$(H, S) \Downarrow_{(q,q')} X$.  It is similar to the judgement
presented in the overview section appart  from the fact that
two resource numbers are tracked instead of one.  Intuitively
$(q, q')$ means that $q$ resources units are necessary to execute
the program and that $q'$ units are available after its execution.
For example, $\code{alloc}(3)$ consumes 3 heap cells, so
the pair resulting from its evaluation would be $(3, 0)$; for
$\code{free}(3)$, the pair would be $(0, 3)$.  When these
statements are sequenced with \code{alloc} first, then \code{free},
the evaluation yields $(3, 3)$.  These resource pairs can
be composed with the following operation
$$
(q_1, q'_1) \cdot (q_2, q'_2) =
\left\{
\begin{array}{ll}
(q_1, q'_2 + q'_1 - q_2) & \mbox{if } q'_1 \ge q_2 \\
(q_1 + q_2 - q'_1, q'_2) & \mbox{if } q'_1 < q_2
\end{array}
\right.
.
$$
This operation can be seen as the sequential composition of
two resource consumptions, note that it is not a commutative
operation, the same way resource freeing and allocation do not
commute.
With $(0, 0)$ as a neutral element, the above composition defines
a monoid structure on $\mathbb Q_0^+ \times \mathbb Q_0^+$.
It is possible to inject $q \in \mathbb Q$ in this monoid with by taking
$(q, 0)$ if $q \ge 0$ and $(0, -q)$ otherwise.  This injection
is used implicitely in the semantics.

@@

\and \Rule{S:Skip}
{ }
{ (H, \code{skip}) \Downarrow_{M_k} H }

\and \Rule{S:Assign}
{ }
{ (H, x \gets e) \Downarrow_{M_s} H[x \mapsto \llbracket e \rrbracket_H] }

\and \Rule{S:Assert}
{ H \models C }
{ (H, \code{assert}(C)) \Downarrow_{M_a} H }

\and \Rule{S:WhileStop}
{ \llbracket C \rrbracket_H = \bot }
{ (H, \code{while}(C)~S) \Downarrow_{M_w^3} H }

\and \Rule{S:WhileAbort}
{  k < \llbracket x - y \rrbracket_H
\\ (H, S) \Downarrow_{(q_1, q'_1)} \circ
}
{ (H, \code{while}(k < x - y)~S) \Downarrow_{M_w^1 \cdot (q_1, q'_1) } \circ }

\and \Rule{S:WhileStep}
{  \llbracket C \rrbracket_H = \top
\\ (H, S) \Downarrow_{(q_1, q'_1)} H'
\\ (H', \code{while}(C)~S) \Downarrow_{(q_2, q'_2)} X
}
{ (H, \code{while}(C)~S) \Downarrow_{M_w^1 \cdot (q_1, q'_1) \cdot M_w^2 \cdot (q_2, q'_2)} X }

\and \Rule{S:IfThen}
{  (H, S_1) \Downarrow_{(q, q')} X
\\ \llbracket C \rrbracket_H = \top
}
{ (H, \code{if}(C)~S_1~\code{else}~S_2) \Downarrow_{M_c^1 \cdot (q, q')} X }

\and \Rule{S:IfElse}
{  (H, S_2) \Downarrow_{(q, q')} X
\\ \llbracket C \rrbracket_H = \bot
}
{ (H, \code{if}(C)~S_1~\code{else}~S_2) \Downarrow_{M_c^2 \cdot (q, q')} X }

\and \Rule{S:Seq}
{  (H, S_1) \Downarrow_{(q_1, q'_1)} H'
\\ (H', S_2) \Downarrow_{(q_2, q'_2)} X
}
{ (H, S_1; S_2) \Downarrow_{M_s^1 \cdot (q_1, q'_1) \cdot M_s^2 \cdot (q_2, q'_2)} X }

\and \Rule{S:SeqAbort}
{ (H, S_1) \Downarrow_{(q,q')} \circ }
{ (H, S_1; S_2) \Downarrow_{(q,q')} \circ }

@@

\begin{figure*}[t!]
 \setlength{\progwidth}{.24\linewidth}
  \centering
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  i=0;
  while (i<n) {
    j=i+1;
    while (j<n) {
      if (*) {
	_tick(1);
	j=j-1; n=n-1;
      }
      j=j+1;
    }
    i=i+1;
  }
   \end{lstlisting}
$|[0, n]| \text{ ticks}$
\\[.7\baselineskip]
      {\bf ex1}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  while (n>0 && m>0) {
    n--; m--;
    while (nondet()) {
      n--; m++;
    }
  }
   \end{lstlisting}
$---$
\\[.7\baselineskip]
      {\bf ex2}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  while (n>0) {
    t = x;
    n=n-1;
    while (n>0) {
      if (*) break;
      n=n-1;
    }
  }
   \end{lstlisting}
$2 |[0, n]|$
\\[.7\baselineskip]
      {\bf ex3}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  flag=1;
  while (flag>0) {
    flag=0;
    while (n>0) {
      n=n-1;       
      flag=1;
    }
  }
   \end{lstlisting}
$2 + 3|[0, n]|$
\\[.7\baselineskip]
      {\bf ex4}
    \end{center}
  \end{minipage}

>>>>

For the sake of simplicity, our quantitative program logic
applies to the Clight language.  This language is the most
abstract intermediate language used by CompCert.  It is very
similar to vanilla C but the expressions are free of
side-effects and there is only one looping construct.
%
\begin{align*}
S &:= \code{assert}~E
\mid \code{skip}
\mid \code{break}
\mid \code{return}~E
\mid E \gets E
\mid E \gets f(E^*)
\\
& \mid \code{loop}~S
\mid \code{if}(E)~S~\code{else}~S
\mid S;S
\end{align*}
%
We added the builtin primitive $\code{assert}$ that checks
if a condition is true before continuing the program
execution.  This will be used to explicit invariants to the
automatic analysis later.

The expressions are left abstract in this presentation.  To
present both the semantics and the program logic, we simply
need an evaluation function that maps any $e \in E$ to
a value $\llbracket e \rrbracket_H$ using $H$ as current
heap.

\section{Semantics}

Values computed by evaluating expressions can be integers, or
memory locations. In our semantics, integers have infinite
precision.  While the program logic would perform as well on
machine integers, our automatic analysis would be unsound.
%
\begin{align*}
V &:= \code{int}~i \mid \code{ptr}~b
\end{align*}

To formally describe the cost of operation of a given
program, our semantics tracks resource consumption while
executing.  To account for all kinds of resources
we leave it as a parameter of the semantics to be specified
by the user.  This parameter, called a metric, is a tuple of rational
numbers with the following form
$$
(
M_a, M_s, M_b, M_r, M_u, M_f,
M_l,
M_c^1, M_c^2,
M_q
)
$$
Each of these rational numbers indicates the amount of resource
consumed by an operation.  If one of these numbers is negative,
it means that resources are released by the program.

The general form of an evaluation judgement is
$(H, S) \Downarrow_{(q,q')} X$.  It is similar to the judgement
presented in the introducing section appart from the fact that
two resource numbers are tracked instead of one.  Intuitively
$(q, q')$ means that $q$ resources units are necessary to execute
the program and that $q'$ units are available after its execution.
For example, $\code{alloc}(3)$ consumes 3 heap cells, so
the pair resulting from its evaluation would be $(3, 0)$; for
$\code{free}(3)$, the pair would be $(0, 3)$.  When these
statements are sequenced with \code{alloc} first, then \code{free},
the evaluation yields $(3, 3)$.  These resource pairs can
be composed with the following operation
$$
(q_1, q'_1) \cdot (q_2, q'_2) =
\left\{
\begin{array}{ll}
(q_1, q'_2 + q'_1 - q_2) & \mbox{if } q'_1 \ge q_2 \\
(q_1 + q_2 - q'_1, q'_2) & \mbox{if } q'_1 < q_2
\end{array}
\right.
.
$$
This operation can be seen as the sequential composition of
two resource consumptions, note that it is not a commutative
operation, the same way resource freeing and allocation do not
commute.
With $(0, 0)$ as a neutral element, the above composition defines
a monoid structure on $\mathbb Q_0^+ \times \mathbb Q_0^+$.
It is possible to inject $q \in \mathbb Q$ in this monoid by
turning it into $(q, 0)$ if $q \ge 0$ and $(0, -q)$ otherwise.
This injection is used implicitely in the semantics.

@@

\and \Rule{S:Assert}
{ {\rm istrue}~\llbracket e \rrbracket_H}
{ (H, \code{assert}~e) \Downarrow_{M_a} \code{seq}~H }

\and \Rule{S:Skip}
{ }
{ (H, \code{skip}) \Downarrow_{M_s} \code{seq}~H }

\and \Rule{S:Break}
{ }
{ (H, \code{break}) \Downarrow_{M_b} \code{brk}~H }

\and \Rule{S:Return}
{ }
{ (H, \code{return}~e) \Downarrow_{M_r} \code{ret}~H~\llbracket e \rrbracket_H }

\and \Rule{S:Update}
{ \llbracket e \rrbracket_H = \code{ptr}~b }
{ (H, e \gets e') \Downarrow_{M_u} \code{seq}~H[b \mapsto \llbracket e' \rrbracket_H] }

\and \Rule{S:FunCall}
{ \Delta~f = ((a_1, \dots, a_n), S)
\\ \llbracket e \rrbracket_H = \code{ptr}~b
\\ (H +\!\!\!+ (a_i \mapsto \llbracket e_i \rrbracket_H)_i, S) \Downarrow_{(q,q')} \code{ret}~H'~v
}
{ (H, e \gets f(e_1, \dots, e_n))
\Downarrow^r_{M_f\cdot(q,q')} \code{seq}~(H'\,\backslash\,(a_i)_i)[b \mapsto v] }

\and \Rule{S:LoopSeq}
{ (H, S) \Downarrow_{(q_1, q_1')} \code{seq}~H'
\\ (H', \code{loop}~S) \Downarrow_{(q_2, q_2')} X }
{ (H, \code{loop}~S) \Downarrow_{(q_1, q_1') \cdot M_l \cdot (q_2, q_2')} X }

\and \Rule{S:LoopBrk}
{ (H, S) \Downarrow_{(q, q')} \code{brk}~H' }
{ (H, \code{loop}~S) \Downarrow_{(q, q')} \code{seq}~H' }

\and \Rule{S:LoopAbort}
{ (H, S) \Downarrow_{(q, q')} X
\\ \forall H.~X \neq \code{seq}~H \land X \neq \code{brk}~H }
{ (H, \code{loop}~S) \Downarrow_{(q, q')} X }

\and \Rule{S:IfThen}
{  (H, S_1) \Downarrow_{(q, q')} X
\\ {\sf istrue}~\llbracket C \rrbracket_H
}
{ (H, \code{if}(C)~S_1~\code{else}~S_2) \Downarrow_{M_c^1 \cdot (q, q')} X }

\and \Rule{S:IfElse}
{  (H, S_2) \Downarrow_{(q, q')} X
\\ {\sf isfalse}~\llbracket C \rrbracket_H
}
{ (H, \code{if}(C)~S_1~\code{else}~S_2) \Downarrow_{M_c^2 \cdot (q, q')} X }

\and \Rule{S:Seq}
{  (H, S_1) \Downarrow_{(q_1, q'_1)} \code{seq}~H'
\\ (H', S_2) \Downarrow_{(q_2, q'_2)} X
}
{ (H, S_1; S_2) \Downarrow_{(q_1, q'_1) \cdot M_q \cdot (q_2, q'_2)} X }

\and \Rule{S:SeqAbort}
{ (H, S_1) \Downarrow_{(q,q')} X
\\ \forall H.~X \neq \code{seq}~H}
{ (H, S_1; S_2) \Downarrow_{(q,q')} X }

@@

\begin{figure*}[t!]
 \setlength{\progwidth}{.24\linewidth}
  \centering
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  i=0;
  while (i<n) {
    j=i+1;
    while (j<n) {
      if (*) {
	_tick(1);
	j=j-1; n=n-1;
      }
      j=j+1;
    }
    i=i+1;
  }
   \end{lstlisting}
$|[0, n]| \text{ ticks}$
\\[.7\baselineskip]
      {\bf ex1}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  while (n>0 && m>0) {
    n--; m--;
    while (nondet()) {
      n--; m++;
    }
  }
   \end{lstlisting}
$---$
\\[.7\baselineskip]
      {\bf ex2}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  while (n>0) {
    t = x;
    n=n-1;
    while (n>0) {
      if (*) break;
      n=n-1;
    }
  }
   \end{lstlisting}
$2 |[0, n]|$
\\[.7\baselineskip]
      {\bf ex3}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  flag=1;
  while (flag>0) {
    flag=0;
    while (n>0 && *) {
      n=n-1;       
      flag=1;
    }
  }
   \end{lstlisting}
$2 + 3|[0, n]|$
\\[.7\baselineskip]
      {\bf ex4}
    \end{center}
  \end{minipage}
